\chapter*{Preface}

There are usually two sides to technology: using it and working on it.
Loosely, I call this driver's ed vs. auto shop. In the former we learn
to operate a car: parking, signaling, turning, minding the rules of the
road. In the latter we would learn to repair the various systems on
a car. Notice I say ``would learn.'' I've never taken auto shop. My
repair skills are limited. This is typical. More people need to know
how to operate a car than need to know how to work on one.

For the expert, shop is more interesting. The temptation to fixate on it
is strong. This definitely holds for books and courses about compilers.
This book takes the driver's ed approach exclusively. It is completely
distinterested in how finite automata and context free grammars work
theoretically and how they are implemented. Our goal will be to use
tools to make languages.

One of the important inovations in software development since the start
of my career has nothing to do with code. It is the
agile process. One of the key concepts of agile development is the Minimum
Viable Product (MVP). The idea is to build something that will run and
deliver some value to the users. After it is in the users' hands, revisit
the product to add features. That will be my approach here.

We will start with the smallest language I could conceive. It will only add
single digit integers. When it does so, it will only print the
result to standard out. Then we will commence adding features to this
language. How far you take it is up to you. At least we want variables,
flow of control (conditional logic and some sort of loop), plus
functions which return values which can be called recursively.
We could go on to add types and type checking, arrays, and many other
features. Remember the goal: to learn the tools of language construction
and play with the language feature trade offs. If you are a student,
when you complete this work, you should be able to imagine implementing
a domain specific language or code generator. You will also learn
the basic concepts in case you have an interest in working on production
compilers (the auto shop part).

The main tool of this course is ANTLR4 which was written by Terence Parr.
He wrote an excellent book about it called
{\it The Definitive ANTLR4 Reference}.
You will probably want to consult that, or the various forms of online
documentation including especially antlr.org. I will shamelessly begin
by implementing the language for arithmetic he lays out there. Extensions
and poor choices are mine alone.

You can use ANTLR4 with various languages including C, C\verb+#+, JavaScript,
and Python. Here I use Python. It is cleaner to look at than some
other languages. If you are going to follow along, choose a language
that is comfortable for you. Follow the insructions to obtain the
run time for your chosen language at antlr.org.

Let's get started.
